VIPE: Virtualized Intermittently-Powered Executor

### 2017-07-23 paper/background.tex
WAR inconsistency section complete

Supplying power to tiny embedded computers using batteries alone is not sustainable. European Commission estimates that more than 160 kilotons of consumer batteries enter the European Union annually~\cite{eu_batteries_2016}. Since not all batteries are recycled, they constitute a potential environmental hazard. A potent solution is to replace batteries with smaller, but environment friendly {(super-)capacitors} (promising millions of charge/discharge cycles~\cite[Sec. I]{ongaro_pwre_2012}) and to supply them {\em directly from the ambient energy sources}---creating a truly self-sustainable computing device. The limitation of the capacitor as storage is its lower than battery energy density, making effective capacitor discharge times orders of magnitude smaller than for a battery.

Given current technology development, battery-less systems are best suited for very long-term sensing and monitoring where access to recharge is either prohibitive or impossible. These include battery-less image capture and processing~\cite{naderiparizi_rfid_2015}, insect monitoring~\cite{thomas_jbcs_2012} or implantable~\cite{rodriguez_tbcs_2015}  and digestible~\cite{nadeau_naturebio_2017} sensors.

@@

\begin{table}
	\begin{tabular}{|c|c|}
	\hline
	Platform name & Storage capacitor size \\
	\hline \hline
	Moo~\cite{moo} & 0.1\,F \\
	WISPCam~\cite{naderiparizi_rfid_2015} & 6.08\,mF \\ %tested [11.24, 17.45, 21.98]\,mF
	NFC-WISP~\cite{zhao_rfid_2015} & 300\,$\mu$F \\
	NeuralWISP~\cite{holleman_biocas_2008} & 100\,$\mu$F \\
	WISP~\cite{wisp5} & 47\,$\mu$F \\
	{\em BioImpedance} sensor~\cite{rodriguez_tbcs_2015} & 20\,$\mu$F \\
	{\em Ingestible} sensor~\cite{nadeau_naturebio_2017} & 220\,$\mu$F\\
	\hline
	\end{tabular} 
\caption{Comparison of {\em default} energy storage sizes for various battery-less platforms. Observe a huge variation in storage capacity. We note that values for other representative platforms~\cite{medusa_farsens_2017,talla_imwut_2017,liu_sigcomm_2013,parks_sigcomm_2014} were not reported in their respective papers.}
\label{table:capacitor}
\end{table}

@@

\begin{figure}
	\centering
	\fbox{\texttt{text}}
	%\includegraphics[width=0.25\columnwidth]{figures/code_demo_incosistency}
	%	for (byte = 0; byte < nBytes; ++byte){\\
	%	checkpoint();\\
		%data = reflectData(message\[byte\]) $^$ (remainder >> (WIDTH - 8));\\
		%remainder = crcTable[data] $^$ (remainder << 8);}
	\caption{Code demonstrating WAR and volatile memory checkpointing.}
	\label{fig:code_demo_incosistency}
\end{figure}

Fig.~\ref{fig:code_demo_incosistency} demonstrates the problem through example.

>>>>

Supplying power to tiny embedded computers using batteries alone is not sustainable. European Commission estimates that more than 160 kilotons of consumer batteries enter the European Union annually~\cite{eu_batteries_2016}. Since not all batteries are recycled, they constitute a potential environmental hazard. A potent solution is to replace batteries with smaller, but environment friendly {(super-)capacitors} (promising millions of charge/discharge cycles~\cite[Sec. I]{ongaro_pwre_2012}) and to supply them {\em directly from the ambient energy sources}---creating a truly self-sustainable computing device. The limitation of the capacitor as a storage device is energy density that is inferior to a battery, making effective capacitor discharge times orders of magnitude smaller than for a battery.

Given current technology development, battery-less systems are best suited for very long-term sensing and monitoring where access to recharge is either prohibitive or impossible. These include battery-less image capture and processing~\cite{naderiparizi_rfid_2015}, animal monitoring~\cite{thomas_jbcs_2012} or implantable~\cite{rodriguez_tbcs_2015} and digestible~\cite{nadeau_naturebio_2017} sensors.

@@

\begin{table}
	\centering
	\begin{tabular}{|c|c|}
	\hline
	Platform name & Storage capacitor size \\
	\hline \hline
	Moo~\cite{moo} & 0.1\,F \\
	WISPCam~\cite{naderiparizi_rfid_2015} & 6.08\,mF \\ %tested [11.24, 17.45, 21.98]\,mF
	NFC-WISP~\cite{zhao_rfid_2015} & 300\,$\mu$F \\
	NeuralWISP~\cite{holleman_biocas_2008} & 100\,$\mu$F \\
	WISP~\cite{wisp5} & 47\,$\mu$F \\
	{\em BioImpedance} sensor~\cite{rodriguez_tbcs_2015} & 20\,$\mu$F \\
	{\em Ingestible} sensor~\cite{nadeau_naturebio_2017} & 220\,$\mu$F\\
	\hline
	\end{tabular} 
\caption{Comparison of {\em default} energy storage sizes for various battery-less platforms. Observe a huge variation in storage capacity. We note that values for other representative platforms~\cite{medusa_farsens_2017,talla_imwut_2017,liu_sigcomm_2013,parks_sigcomm_2014} were not reported in their respective papers.}
\label{table:capacitor}
\end{table}

@@

\begin{figure}
	\centering
	\subfloat[Simplified C code snippet of a CRC calculation from~\cite{hicks_mibench2_2016}: per-byte message division by a polynomial; \texttt{NV} denotes non-volatile variable declaration]{\includegraphics[width=\columnwidth]{figures/crc_example}\label{fig:crc_example}}\\
	\subfloat[Consecutive execution steps of the loop body in the snippet above: non-volatile checkpointing did not guarantee data consistency as data has been manipulated (line 10) with stale reminder (line 3)]{\includegraphics[width=\columnwidth]{figures/crc_example_war}\label{fig:crc_example_war}}
	\caption{Code example demonstrating effect of write after read on volatile memory checkpointing.}
	\label{fig:code_demo_incosistency}
\end{figure}

Nevertheless, regular checkpointing of program state to a non-volatile memory still does not guarantee consistency. This phenomenon has been discussed extensively in e.g.~\cite{chain,alpaca}. For the completeness of this exposition we will provide our own example demonstrating this problem, please refer to Fig.~\ref{fig:code_demo_incosistency} and compare with~\cite[Sec 2.3]{alpaca},~\cite[Sec. 2.1]{chain}. We take an example of a fast procedure to calculate cyclic redundancy check (CRC) from~\cite{hicks_mibench2_2016}. Therein a CRC must be calculated for $\texttt{nBytes}$-long message $\texttt{msg}$ with specified initial reminder $\texttt{Rmnd}$ and constant, $\texttt{Rmnd}$ variable type-dependent, $\texttt{Wdth}$. Despite regular, fixed position checkpoints (line 3 in Fig.~\ref{fig:crc_example}) of all variables which are declared as non-volatile (denoted as \texttt{NV}) variable $\texttt{data}$ is calculated incorrectly. This is because of a write after read (WAR) hazard of variable $\texttt{Rmnd}$. WAR hazard is caused by consecutive write and read in-between two consecutive checkpoints using $\texttt{checkpoint()}$ function: $\texttt{data}$ variable is written \emph{before} (line 10 in Fig.~\ref{fig:crc_example}) $\texttt{Rmnd}$ is read (line 6 in Fig.~\ref{fig:crc_example}), or more specifically---updated before a new execution loop.
